---
title: "Using basic_summaries()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using basic_summaries()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bsu.tools)
```

The `basic_summaries()` function produces a combined summary table and set of
figures for a data frame.  It wraps `gtsummary::tbl_summary()` and returns an
S7 object (`basic_summaries_result`) with three properties:

- **`tbl_gt`** – a `gt` table with small inline figures embedded in a
  *Figures* column.
- **`raw_table`** – the underlying `gtsummary` table object, useful for
  further customisation.
- **`figures`** – a named list of `ggplot2` objects (one per summarised
  variable).

## Without a grouping variable

When `by_var` is `NULL` (the default), numeric variables are shown as
histograms and categorical variables as bar charts.

### Example: `mtcars`

```{r mtcars-no-group, eval=FALSE}
result_mtcars <- basic_summaries(
  data = mtcars,
  vars = c("mpg", "hp", "wt", "cyl")
)
result_mtcars
```

The printed output is the `gt` table. You can also access each component
directly:

```{r mtcars-components, eval=FALSE}
# raw gtsummary table
result_mtcars@raw_table

# individual ggplot figures
result_mtcars@figures[["mpg"]]
result_mtcars@figures[["cyl"]]
```

## With a grouping variable

When `by_var` is supplied, numeric variables are shown as box plots (one box
per group) and categorical variables as dodged bar charts.  An *Overall*
column is also added automatically.

### Example: `iris` grouped by `Species`

```{r iris-by-species, eval=FALSE}
result_iris <- basic_summaries(
  data = iris,
  vars = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
  by_var = "Species"
)
result_iris
```

Access the figures for any variable:

```{r iris-figures, eval=FALSE}
result_iris@figures[["Sepal.Length"]]
result_iris@figures[["Petal.Width"]]
```

## Selecting a subset of variables

You can pass any subset of column names to `vars`.  Variables not listed are
excluded from both the table and the figures.

```{r subset, eval=FALSE}
basic_summaries(
  data = mtcars,
  vars = c("mpg", "am")
)
```
