---
title: "Using git within RStudio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using git within RStudio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette walks through using [git](https://git-scm.com/) with
[RStudio](https://posit.co/download/rstudio-desktop/). git is a version control
system that tracks changes to your project over time, and it pairs naturally
with the collaborative workflow used across the BSU.

Most of the guidance here follows the excellent, free resource
[Happy Git and GitHub for the useR](https://happygitwithr.com/) by Jenny Bryan.
When you want more detail than this vignette provides, that is the place to go.

The code chunks below are shown for reference only and are not run when the
vignette is built. Shell chunks are meant to be typed at a terminal (the
RStudio **Terminal** tab), while R chunks are meant for the RStudio
**Console**.

## 1. Install git

Before you can use git you need to install it on your computer. Download and
install git from <https://git-scm.com/downloads>.

- **Windows:** Install [Git for Windows](https://gitforwindows.org/). This also
  gives you the "Git Bash" shell.
- **macOS:** The easiest route is to open the Terminal and run `git --version`.
  If git is not already installed, macOS will offer to install the Command Line
  Tools for you.
- **Linux:** Use your distribution's package manager, e.g.
  `sudo apt-get install git` (Debian/Ubuntu) or `sudo dnf install git`
  (Fedora).

Confirm the installation by opening the **Terminal** tab in RStudio and running:

```bash
git --version
```

If you see a version number, git is installed. See
[Happy Git chapter 6](https://happygitwithr.com/install-git.html) for more
detail.

You will also need a free account on [GitHub](https://github.com/), since the
BSU hosts its repositories there.

## 2. Introduce yourself to git

git records *who* made each change, so you need to tell it your name and email.
Use the same email address that is associated with your GitHub account.

The `usethis` package (already a dependency of `bsu.tools`) makes this easy from
the RStudio **Console**:

```{r}
usethis::use_git_config(
  user.name = "Jane Doe",
  user.email = "jdoe@example.com"
)
```

Equivalently, you can do this from the **Terminal**:

```bash
git config --global user.name "Jane Doe"
git config --global user.email "jdoe@example.com"
```

Check your settings at any time with:

```{r}
usethis::git_sitrep()
```

To connect to GitHub without typing your password repeatedly, set up a
**Personal Access Token (PAT)**. Generate one and store it with:

```{r}
usethis::create_github_token()   # opens GitHub to create a PAT
gitcreds::gitcreds_set()         # paste the PAT when prompted
```

See [Happy Git chapter 7](https://happygitwithr.com/hello-git.html) and the
[credential caching chapter](https://happygitwithr.com/https-pat.html) for
details.

## 3. Clone a repository

*Cloning* makes a local copy of a repository that is already on GitHub. In
RStudio the smoothest path is to create a new **Project** from version control:

1. **File > New Project > Version Control > Git**.
2. Paste the repository URL, e.g.
   `https://github.com/ucdavis/bsu.tools.git`. Find this under the green
   **Code** button on the GitHub repository page.
3. Choose a local folder for the project and click **Create Project**.

RStudio will clone the repository and open it as a project, complete with a
**Git** pane in the upper-right.

You can accomplish the same thing from the **Console** with `usethis`:

```{r}
usethis::create_from_github(
  "ucdavis/bsu.tools",
  destdir = "~/projects"
)
```

See [Happy Git chapter 15](https://happygitwithr.com/existing-github-first.html)
for the "GitHub first" workflow.

## 4. Using branches

A *branch* lets you work on a change in isolation without affecting the main
line of development (usually called `main`). The BSU workflow is to create a
branch for each new feature or fix, then merge it back via a Pull Request.

Create and switch to a new branch from the **Console**:

```{r}
usethis::pr_init("my-new-feature")
```

Or from the **Terminal**:

```bash
git checkout -b my-new-feature
```

In the RStudio **Git** pane, the current branch is shown in the top-right; you
can switch between existing branches using the branch dropdown. Keep branch
names short and descriptive.

See [Happy Git chapter 22](https://happygitwithr.com/git-branches.html) for more
on branching.

## 5. Basics of commit / push / pull

Once you have made changes, the core loop is **commit**, **push**, and **pull**.

- **Commit** records a snapshot of your changes in the local repository.
- **Push** sends your local commits up to GitHub.
- **Pull** brings other people's commits down from GitHub to your computer.

### Committing in RStudio

1. In the **Git** pane, check the boxes next to the files you want to include
   (this is *staging*).
2. Click **Commit**.
3. Write a short, meaningful commit message describing *what* changed and
   *why*, then click **Commit** again.

The equivalent **Terminal** commands are:

```bash
git add .
git commit -m "Add a helpful summary of what changed"
```

### Pushing and pulling in RStudio

Use the green **up arrow** (Push) and blue **down arrow** (Pull) in the **Git**
pane. As a rule of thumb, **pull before you start working** and **push when you
reach a good stopping point**.

The equivalent **Terminal** commands are:

```bash
git pull
git push
```

See [Happy Git chapter 21](https://happygitwithr.com/git-basics.html) for a
walk-through of the commit/push/pull cycle.

## 6. Basics of Pull Requests

A *Pull Request* (PR) proposes merging the changes from your branch into
`main`, and gives collaborators a place to review and discuss them before they
are merged.

The `usethis` package streamlines the whole cycle:

```{r}
# Start a branch for your work
usethis::pr_init("my-new-feature")

# ... edit files, then commit them via the Git pane or the Terminal ...

# Push the branch and open a PR in your browser
usethis::pr_push()
```

`pr_push()` pushes your branch to GitHub and opens the page where you fill in
the PR title and description, then click **Create pull request**.

After the PR is reviewed and merged on GitHub, tidy up locally with:

```{r}
usethis::pr_finish()
```

You can also open a PR manually on the GitHub website: after pushing your
branch, GitHub shows a **Compare & pull request** button.

See the Happy Git
[Pull request helpers chapter](https://happygitwithr.com/pr-functions.html) for
the full `usethis` Pull Request workflow.

## Where to go next

This vignette is only a starting point. For a thorough, R-focused treatment of
git and GitHub, work through
[Happy Git and GitHub for the useR](https://happygitwithr.com/).
